Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-11-29 responseToReviews.md

Page 12, better to state whether attention effects were consistent with response gain or contrast gain to make the result more concrete for the reader.

> Reynolds and Heeger (and Boynton) point out that the range of different types of change reported in the literature (response, contrast or mixtures of both) is explained by subtle differences in stimulus configuration (for example, the size and location of the 'probe' and any masking components relative to the size of the 'attentional spotlight'). We now note that the underlying model is one of contrast gain control (multiplicative amplification of the input rather than the output).

Discussion of time varying response paragraph on page 12 is not that relevant to contrast response function measurement, nor is the BCI section.  These applications of SSVEP and could be deleted or the relationship to contrast response function made clearer.

> Originally, we envisaged this document as a primer on SSVEP contrast measurements for those coming to the field for the first time. In particular, we were struck by the large number of recent papers that use multi-frequency SSVEP contrast flicker as an input to BCI systems. There have been at least 400 such publications ('SSVEP AND BCI') in the last 5 years alone (Kcompared to around 10 for amblyopia ('SSVEP AND amblyopia')and 80 for ('SSVEP AND clinical'). Some of these publications measure dynamic changes in covert spatial attention, some measure changes in featural attention, some measure fixation). We know from personal experience that many of the researchers using contrast flicker as a BCI input would benefit from a slightly more applied guide such as this and the BCI section adds very little to the overall length of the manuscipt at this point.

### 2024-11-29 responseToReviews.md

Two-limbed contrast response functions. Campbell and Maffei (J. Physiol,1970) in the earliest measurements of the VEP contrast response function reported that it sometimes had two “limbs” – one at low contrast and a second one with a steeper slope at higher contrasts.  Two-limbed contrast response functions have also been found by Nakayama and Mackeben (Vision Res,1982) in macaque with electrodes on the dura, by Apkarian and Tyler (Vision Res, 1985),  Norcia et al., (Vision Res,1989) and Souza et al., (IOVS, 2007), to name a few.  As noted, the sigmoidal model assumes that the population response is unitary, but the two-limbed functions suggest that it is a mixture of populations. The two limbs may not be apparent when the function is sampled with only a few points over a wide range of contrast.

Harmonics with different CRFs.  Bobak et al., (IOVS, 1986), found that the first harmonic (1f) and second harmonic (2f) of the pattern onset/offset SSVEP had different thresholds but similar slopes once above threshold. Norcia et al., (TVST, 2024) found different thresholds and slope for 1f and 2f responses driven by contrast modulated dynamic noise. This again suggests the presence of multiple underlying populations.  The behavior also raises the question of whether to pool harmonics when estimating the response function. This was alluded to in the simulation, but could be expanded on.

The lack of dynamics in the normalization model, as presented should be discussed, the brief mention of Tsai et al. aside. The authors may wish to cite another framework that explicitly models dynamics (e.g. amplitude and phase) in the context of the contrast response function (Zemon and Gordon, Vision Res, 2006) .

### Response
> We thank the reviewer for these positive comments. In general this was intended to be an overview in the sense of a 'primer' for researchers coming to the SSVEP for the first time rather than a detailed historical account of the field - hence the inclusion of code examples and relatively domain-agnostic language. However, the reviewer raises some important general issues and we are happy to address them: Specifically, we include references to the idea that SSVEP can engage more than one type of neuron, that the population responses is not trivially derived from what is known about the properties of individual neurons and that fitting with a sigmoidal response function may mask some complexities in the CRF that are more readily seen with denser sampling. We think the Bobak reference is IOVS 1984 and we have also included the Kaestner TVST 2024 reference (thank you for drawing our attention to both of these).

@@

Page 5 bottom. The reciprocal effect of test and masker was first shown by Regan and Regan  (1988; cited)  for flicker modulation depth and then for gratings by Brown et al., (IOVS,  1999 for dichoptic masking) and  then  by candy et al., (J. Neurosci, 2001) along with Busse et al. (Neuron, 2009) as cited.

> Thanks - we have included these references

 It’s good to have cited Regan and Regan paper that introduced the notion of fingerprints in the spectrum being tied to the nature of the non-linearity. Maybe expand their “fingerprint” approach a little more.

> We have expanded a little but are also conscious of limitations to this approach (see also the discussion re unique inputs and system order below). We think there might be some ways to relax these constraints a little by incorporating information about likely transducer functions at different stages (essentially fitting parameters rather than performing a full system identification) but this still requires many inputs.

@@

Page, 14. It’s not clear from the text how one would study feedback with the SSVEP, please elaborate.

> In the paper we write that we would look for modulation of subcortical responses due to top-down signals. The most natural things to look for might be interocular normalization (because eye layers are segregated in the LGN) and attentional gain control (because covert spatial attentional modulation must arise in cortex). This the type of feedback we are thinking of - not cortico-cortico feedback (which is hard to isolate with SSVEP)

In the last section, it would be worth mentioning that in order to “identify” a non-linear system, one needs multiple inputs. If the system is 2nd order, a minimum of two inputs is needed. If the system non-linearity is higher order than that, then more simultaneous inputs are needed (see Boyd, Tang and Chua 1983, IEEE Trans Circuits and Systems  and  Chua and Liao, 1991, Int. J. Circuit Theory and Applications).

> This is a good point and we now include the Boyd reference in the section on fingerprinting.

@@

## Reviewer: 2

# Comments to the Author
This manuscript is intended to be a review of the field of steady-state visual evoked potentials (ssVEPs) over the past 70 years.  The authors state in the introductory paragraph “Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.” The material included and the references cited, however, are quite limited and omit seminal work in this area.

> We thank the reviewer for their helpful comments. As we noted above, our intention here was to provide an introduction to the SSVEP technique with appropriate simple code samples that would allow relative newcomers to the field to implement the the technique. We are, however, keen to avoid misrepresenting the history of the technique and are happy to include appropriate references.

1. All abbreviations should be defined upon first use.

> Done

2. The following statement is unclear and inaccurate: “On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.” The term “on-off flicker” is typically used for simple luminance flicker without spatial pattern contrast. In the context of pattern stimulation, “on-off” can refer to “appearance-disappearance” of a spatial pattern with its contrast modulated in time. When the pattern appears and disappears into a uniform field of equal space-average luminance, the light and dark elements are modulated in counterphase. Any linear contribution to the response elicited by each set of elements (e.g., bars or checks) is expected to be canceled when summed at the electrode site on the scalp. Thus, the response at 1F to this stimulus is in fact a nonlinear response.

> Thank you for this explanation. We have now reworded the explanation to make this clearer. We did take care to define what we mean by 'on-off' flicker in the preceding sentence but we now clarify the alternative 'appearence/disappearence' terminology. The point we were making was that odd harmonic responses (1F) are a feature of this this type of modulation because the responses to the appearance and disappearance of a pattern (or mean field) need not be completely symmetrical (See Zemon, et al 1988 for an explicit demonstration).

This type of stimulation does not separate contributions from “independent populations of on- and off-cells.”   There are other ssVEP techniques developed that do tap ON and OFF pathways selectively, which are not cited here. Contrast-reversal stimuli which involve symmetrical modulation of light and dark elements cancel odd harmonic frequency components in the response, however, both kinds of stimulus techniques mentioned here elicit only nonlinear responses.

> We have clearly not been sufficiently clear here. We do not think we used the term 'separation' and we agree that a) there are infrequently-used paradigms that can target ON and OFF pathways more effectively (for example, 'sawtooth' temporal modulations that provide asymmetrical onset or offset edges) and b) responses from spatial features of differing polarity will tend to be combined at the scalp as long as the spatial frequency is high enough. We have expanded this section to attempt to make the general point clear.

3. The statement “SSVEPs have proven to be an excellent measure of early chromatic processing as well” again fails to cite seminal work in this area.

> We have included three earlier papers by Regan on this topic.

4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).

5. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?

6. The authors use the term “frequency tagging” which has been used in the literature to refer to stimulating different regions of the stimulus field or fellow eyes with different sinusoidal frequencies and measuring the intermodulation (sum and difference) frequency responses that result from nonlinear interactions. They do not, however, cite the foundational work in this field.

7. There is a large body of work that used ssVEPs to investigate contrast processing developmentally (infants through adulthood) and in various disease states, which is mostly missed in this review article.

8. Figures should be labeled.
Thus, the stated intention of this manuscript (“Here we describe how SSVEPs have been used to study visual contrast over the past 70 years”) fails to be met. Only select articles from a few sources are covered, and therefore, the richness and value of this technique is lost.

### 2024-11-29 review.qmd
Now onto the other reviewer. Losing the will to live...

---
title: "Measuring contrast processing in the visual system using SSVEP"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  #docx: 
    #omitFigures: true
  pdf: default
    #keep-tex: true

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR, and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965, @schadeOpticalPhotoelectricAnalog1956], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2005; @Di-Russo2007]. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while Kaester et al [@KaestnerTVST2024] show both different CRF slopes and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@McKeefry1996; @Di-Russo2001]   (see also [@Baseler1997]).

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing (e.g. [@Zemon2006Dynamic]) and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, [@Tsai2012] demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

@@

Attention exerts a profound influence on visual performance: for example, instructing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI BOLD responses 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006; @Muller2000; @Morgan1996; @Muller1998].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces, which seek to control some aspect of a computer using neural signals. The high SNR and precise frequency resolution of SSVEPs are advantages for this approach. Typical studies may involve presenting an array of stimuli at different flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010], the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is generally present, this modulation and the associated changes in gain control will affect the entire pattern of self- and intermodulation terms, allowing the choice to be decoded by a multivariate pattern classifier.

>>>>

---
title: "Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  #docx: 
    #omitFigures: true
  pdf: default
    #keep-tex: true

@@

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, clinical conditions, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for over a century. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged' (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high signal to noise ratio (SNR), and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years [@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This paper can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to changes in cone photoreceptor activations across time and space. This property, referred to as 'contrast', sets the fundamental limits of our visual abilities, which remain steady over a remarkably wide range of environmental light levels. The human response to contrast can be studied using many different techniques. Early work used psychophysical methods to measure contrast sensitivity [@Campbell1965, @schadeOpticalPhotoelectricAnalog1956], defined as the inverse of the lowest contrast that can be reliably detected. But neural responses can also be measured more directly using techniques such as functional magnetic resonance imaging (fMRI), magnetoencephalography (MEG), and electroencephalography (EEG). Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual visually-evoked potentials (VEPs) reveals multiple generators throughout visual cortex [@Di-Russo2005; @Di-Russo2007]. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies.

Two common stimulus variants involve sinusoidal on-off (or 'appearance/disappearance') flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can excite independent populations of on- and off-cells in the visual system once per cycle. For spatial patterns, the contributions of individual cortical cells to these excitations are thought to sum at the scalp and generate an average of their onset responses on each cycle. For very low spatial frequencies (including zero) resposes at the scalp will be dominated by a single on- or off- cell type on each half cycle.

If the amplitudes and waveforms of responses to the stimulus appearance and disappearnce are not perfectly balanced, a response at the fundamental flicker frequency, known as 1F, and its integer harmonics can be produced. For stimuli of any spatial frequency (including zero), an imbalance can be caused by differences in either the onset- and offset- responses of either on- or off-cells. For example, if only on-sensitive neurons were present (firing to the onset, but not the offset of local increases in luminance) the response to the on/off (or 'appearance/disappearance') presentation of a grating of any spatial frequency below the resolution limit would consist entirely of odd harmonics.

By comparison, counterphase flickering patterns generate two essentially identical transients per cycle and therefore does not produce a response at 1F, only even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker (either on/off or contrast reversig) tends to produce additional spectral harmonics compared to sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while Kaester et al [@KaestnerTVST2024] show both different response slopes and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973, @reganColourCodingPattern1975, @reganEffectStimulusColour1966a, @McKeefry1996; @Di-Russo2001] (see also [@Baseler1997]).

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the lateral geniculate nucleus (LGN) [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, and a full system identification may require the use of multiple different input frequencies [@boydMeasuringVolterraKernels1983], careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing (e.g. [@Zemon2006Dynamic]) and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, [@Tsai2012] demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. More recent work has used this approach to study binocular integration in both normal subjects and amblyopes[@Hou-houchuan2021,@Hou2020]. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

@@

Attention exerts a profound influence on visual performance: for example, instructing people to attend covertly to a spatial location improves their performance on a target detection task significantly [@Posner1980; @Cameron2002; @Carrasco2000; @Bashinski1980; @Morrone2004; @Pestilli2009]. In principle, this enhancement may be driven both by modulation of the underlying signal or noise characteristics, or by additional decision-theoretic factors such as reduction in spatial uncertainty [@Gould2007; @Petrov2006]. Early experiments with non-human primates showed little evidence for attentionally-driven changes in neuronal spike rates [@Luck1997; @Marcus2002; @McAdams1999; @Mehta2000a; @Mehta2000b], but with the advent of spatially-resolved human brain imaging methods in the late 1990s it became apparent that spatial attention was linked to frank changes in both fMRI 
[@Tootell1998; @Brefczynski1999; @Buracas2007; @Gandhi1999; @Kastner1999; @Li2008; @Murray2008; @Silver2007; @Somers1999] and EEG signals [@Ding2006; @Muller2000; @Morgan1996; @Muller1998].

@@

One widespread recent application of the SSVEP technique is in the design of brain-computer interfaces (BCIs), which seek to control some aspect of a computer using neural signals. The high SNR and precise frequency resolution of SSVEPs are advantages for this approach. Typical studies may involve presenting an array of stimuli at different flicker frequencies, and having the participant select one either by overt attention (i.e. shifting fixation to foveate the selected stimulus) or covert attention (i.e. deploying attention to one stimulus whilst keeping fixated) [@Middendorf2000]. Because SSVEP signals are highly sensitive to both visual field position [@Ales2010; @Di-Russo2007] and attentional state [@Morgan1996; @Muller1998; @Verghese2012; @Lauritzen2010], the response to the selected stimulus will typically increase relative to the others, allowing it to be identified by an on-line algorithm. Because more than one stimulus frequency is generally present, this modulation and the associated changes in gain control will affect the entire pattern of self- and intermodulation terms, allowing the choice to be decoded by a multivariate pattern classifier.
